---
title: "Examining NCAA Play By Play Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}
library(tidyverse)
library(scales)

theme_set(theme_minimal())


play_by_play <- read_csv("PlayByPlay_2019/Events_2019.csv")
players_2019 <- read_csv("PlayByPlay_2019/Players_2019.csv")

```


What events are there in the dataset, and how many of each are there?

```{r}
play_by_play %>% count(EventType, sort = T) %>% DT::datatable()

# get the list of event types to see which would relate to shooting 
play_by_play %>% count(EventType) %>% pull(EventType)

```


Classifying the different shooting variables. This will include layups, dunks, 2-point jumpers, threes and free-throws

```{r}
# there are some shooting variables that can probably be condensed - tip ins and dunks
paint_attempts_made <- c("made2_dunk", "made2_lay", "made2_tip") 
paint_attempts_missed <- c("miss2_dunk", "miss2_lay", "miss2_tip") 
paint_attempts <- c(paint_attempts_made, paint_attempts_missed)

# create variables for field goals made, and also field goals attempted (which includes the sum of FGs made and FGs missed)
FGM <- c("made2_dunk", "made2_jump", "made2_lay",  "made2_tip",  "made3_jump")
FGA <- c(FGM, "miss2_dunk", "miss2_jump" ,"miss2_lay",  "miss2_tip",  "miss3_jump")

# variable for three-pointers
ThreePointer <- c("made3_jump", "miss3_jump")

#  Two point jumper
TwoPointJump <- c("miss2_jump", "made2_jump")

# Free Throws
FT <- c("miss1_free", "made1_free")

# create a variable in the df for whether the attempts was made or missed
play_by_play <- play_by_play %>% 
  mutate(shot_outcome = ifelse(grepl("made", EventType), "Made", ifelse(grepl("miss", EventType), "Missed", NA))) %>% 
  mutate(FGVariable = ifelse(EventType %in% FGA, "Yes", "No"),
         AttemptType = ifelse(EventType %in% paint_attempts, "PaintVariable", ifelse(EventType %in% ThreePointer, "ThreeVariable", ifelse(EventType %in% TwoPointJump, "TwoPointJumper", ifelse(EventType %in% FT, "FreeThrow", "NoAttempt")))))

```


There were `r comma(sum(play_by_play$FGVariable == "Yes"))` field goals recorded in the data set provided. Looking at the NCAA site (https://stats.ncaa.org/rankings/conference_trends), there were actually almost 676K field goal attempts, but the percentages appear very similar.


```{r}

play_by_play %>% 
  group_by(AttemptType) %>% 
  summarise(n_shots = n()) %>% 
  filter(AttemptType != "NoAttempt") %>% 
  ggplot(aes(x= AttemptType, y= n_shots)) +
  geom_col(fill = "steelblue", colour = "grey") +
  geom_text(aes(label = comma(n_shots)), hjust =  1, colour = "white") +
  scale_y_continuous(labels = comma, name = "Number of Attempts") +
  coord_flip() +
  ggtitle("MORE THREES ATTAMPTED THAN JUMPERS INSIDE THE LINE") +
  theme(axis.title.y = element_blank(), panel.grid.minor = element_blank(), panel.grid.major = element_line(linetype = 2))

```

Threes were the most frequent shot types in the game during the 18/19 season, with `r percent(mean(play_by_play$AttemptType[play_by_play$FGVariable == "Yes"] == "ThreeVariable"))` of all field goals taken being three pointers. Points in the paint (layups, tip-ins and dunks) closely followed, with `r percent(mean(play_by_play$AttemptType[play_by_play$FGVariable == "Yes"] == "PaintVariable"))` of attempts coming from this valuable area. Two point jumpers made up the remainder, with `r percent(mean(play_by_play$AttemptType[play_by_play$FGVariable == "Yes"] == "TwoPointJumper"))` of the attempts.


## NCAA overall shooting averages

```{r}
play_by_play %>% 
  filter(FGVariable == "Yes") %>% 
  group_by(AttemptType, shot_outcome) %>% 
  summarise(n_attempts = n()) %>% 
  mutate(ShootingPercentage = percent(n_attempts / sum(n_attempts))) %>% 
  filter(shot_outcome == "Made") %>% select(-shot_outcome) %>% DT::datatable()

```

The FT% in the games analysed was a touch over 70% - not a bad clip, however lower than the NBA average for the 18-19 season (which was just over 76%).

The overall FG% percentage was 44% on over 628k attempts. When we break this down further, it can be seen that shots in the paint (layups, dunks and tip-ins) were successful on 60% of the 224k attempts, while on two-point jumpers, the percentage fell to 36.4% on over 161k attempts. Suprisingly, tries on threes were only 2% less successful than 2-point jumpers, yet the potential payoff is 50% more points for taking the three. That's what the stat heads keep banging on about anyway.


### Expected points

We can measure how many points each shot attempts expected to yield. To do that, I'll extract the elements of the `PaintVariable` and measure their percentages and shot values.

```{r}
shot_values <- play_by_play %>% 
  filter(FGVariable == "Yes") %>% 
  mutate(EventType = str_remove(EventType, "made"), EventType = str_remove(EventType, "miss")) %>% 
  group_by(EventType, shot_outcome) %>% 
  summarise(n_attempts = n()) %>% 
  mutate(ShootingPercentage = n_attempts / sum(n_attempts),
         n_attempts = sum(n_attempts)) %>% 
  filter(shot_outcome == "Made") %>% select(-shot_outcome)

shot_values %>% mutate(ShootingPercentage = percent(ShootingPercentage)) %>%  DT::datatable()
```


Dunks are by far the most efficient shot type, with the league-wide shooting percentage of 88.4%, yielding an expected points of 1.77 points, while tip-ins are made at a rate of almost 70% worth 1.4 points. It's pretty obvious to see why basketball analytics junkies are screaming from the rooftops for players to stop shooting jumpers from inside the three point line - these shot types have an expected points 0.73 points per attempt...even less than the expected 1.03 points for threes!

```{r}
shot_values %>% 
  mutate(Point = as.numeric(str_extract(EventType, "[[:digit:]]"))) %>% 
  mutate(ExpectedPoints = ShootingPercentage * Point) %>% 
  ggplot(aes(x=reorder(EventType, ExpectedPoints), y= ExpectedPoints)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(ExpectedPoints, 2)), vjust =  1.2, colour = "white") +
  ggtitle("TWO POINT JUMPERS THE LEAST EFFICIENT SHOT TYPE") +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(), panel.grid.major.x = element_line(linetype = 2),
        axis.title = element_blank(), axis.text.y = element_blank(), axis.text.x = element_text(size = 12, angle = 45))
```





## Assists vs Solo Attempts

```{r}

# Rework DF so only shots are included and whatever lead to the shot --------

shooting_analysis <- play_by_play %>% 
  mutate(GameID = paste(Season, DayNum, WTeamID, LTeamID, sep = "_")) %>% 
  group_by(GameID, ElapsedSeconds) %>% 
  mutate(EventType2 = lag(EventType),
         EventPlayerID2 = lag(EventPlayerID))


shooting_analysis <- shooting_analysis %>% 
  mutate(FGVariableAny = ifelse(EventType %in% FGA | EventType2 %in% FGA, "Yes", "No")) %>% 
  filter(FGVariableAny == "Yes") 


# create a variable for if the shot was made, but then the second event was also a made shot
shooting_analysis <- shooting_analysis %>% 
  mutate(Alert = ifelse(EventType %in% FGM & EventType2 %in% FGM, "Alert", "OK"))

shooting_analysis$EventType2[is.na(shooting_analysis$EventType2)] <- "no_second_event"


# create a variable for if there was an assist on the FGM:
shooting_analysis <- shooting_analysis %>% 
  mutate(AssistedFGM = ifelse(EventType %in% FGM & EventType2 == "assist", "Assisted", 
                              ifelse(EventType %in% FGM & EventType2 != "assist", "Solo", 
                                     ifelse(EventType %in% FGM & EventType2 == "no_second_event", "Solo", "None"))))


test <- shooting_analysis %>% 
  filter(Alert == "OK") %>% 
  select(GameID, ElapsedSeconds, EventPlayerID, FGVariable, AssistedFGM) %>% 
  filter(FGVariable == "Yes")


# # add player names
# shooting_analysis <- shooting_analysis %>% 
#   left_join(players_2019 %>% select(PlayerID, PlayerName), by = c("EventPlayerID" = "PlayerID")) %>% 
#   left_join(players_2019 %>% select(PlayerID, PlayerName), by = c("EventPlayerID2" = "PlayerID"))

```


Field goals can be made either assisted, or un-assisted. Players that can create their own shots will usually see their skills transfer somewhat easier in different systems, while players (spot-up shooters) who require a facilitator will usually find their shot success closely tied to their facilitator and system. Lose a good point guard, and a great shooter can come back to the pack fairly quickly. It's the reason why NBA players like Steph Curry and James Harden are so valuable in today's NBA - great offensive players capable of creating for themselves.

It can be seen below that the 2018/19 Div 1 season saw just under 20% of threes un-assisted. Contrast this with two point jumpers, which were un-assisted on 71% of these made attempts. As expected, almost all tip-ins were un-assisted (I don't know how a tip in could be assisted anyway).



```{r}
shooting_analysis %>% 
  filter(AssistedFGM != "None") %>% 
  group_by(EventType, AssistedFGM) %>% 
  summarise(n = n()) %>% 
  mutate(perc = n / sum(n)) %>% 
  ggplot(aes(x=EventType, y= perc, fill = AssistedFGM)) +
  geom_col(colour = "black") +
  geom_text(aes(label = ifelse(AssistedFGM == "Solo", percent(perc), "")), hjust = 1.2, colour = "white") +
  scale_fill_manual(values = c("grey", "steelblue")) +
  scale_y_continuous(labels = percent, name = "Percent Made Shots") +
  coord_flip() +
  ggtitle("SOME SHOT TYPES ARE ASSISTED MORE THAN OTHERS") +
  theme(legend.position = "top", legend.title = element_blank(), axis.title.x = element_text(hjust = 1), axis.title.y = element_blank())
```



# Best Own FG Creators

On average, players took 137 shots for the season.

```{r}
shot_attempts <- play_by_play %>% 
  filter(FGVariable == "Yes") %>% 
  group_by(EventPlayerID) %>% 
  summarise(n_attempts = n()) %>% ungroup() %>% arrange(desc(n_attempts)) %>% left_join(players_2019, by = c("EventPlayerID" = "PlayerID")) 


shot_attempts %>%
  ggplot(aes(x= n_attempts)) +
  geom_histogram(fill = "steelblue", colour = "grey") +
  geom_vline(xintercept = mean(shot_attempts$n_attempts), linetype = 2) +
  annotate("text", x=300, y= 400, label = paste0("Average player\ntook ", round(mean(shot_attempts$n_attempts)), " shots")) +
  labs(x= "Shot Attempts", y= "", title = "PLAYER SHOT ATTEMPTS") +
  theme(axis.title.x = element_text(hjust = 1, size = 12))
```


```{r}
players_over200_shots <- shot_attempts %>% 
  
```





```{r}
league_averages <- shooting_analysis %>% 
  filter(AssistedFGM != "None") %>% 
  ggplot(aes(x=EventType, fill = AssistedFGM)) +
  geom_bar(stat = "count", position = "fill", colour = "black") +
  scale_fill_manual(values = c("grey", "steelblue")) +
  scale_y_continuous(labels = percent) +
  coord_flip() +
  ggtitle("NCAA AVERAGES")


ja_morrant <- shooting_analysis %>% 
  filter(AssistedFGM != "None") %>% 
  filter(EventPlayerID == 656216) %>% 
  ggplot(aes(x=EventType, fill = AssistedFGM)) +
  geom_bar(stat = "count", position = "fill", colour = "black") +
  scale_fill_manual(values = c("grey", "steelblue")) +
  scale_y_continuous(labels = percent) +
  coord_flip() +
  ggtitle("JA MORRANT GETS HIS OWN BUCKETS")


gridExtra::grid.arrange(league_averages, ja_morrant)





shooting_analysis %>% 
  filter(AssistedFGM != "None") %>%
  group_by(shot_outcome, AssistedFGM) %>% 
  summarise(n = n())




threes_made <- shooting_analysis %>% 
  filter(EventType == 'made3_jump') %>% 
  group_by(EventPlayerID, AssistedFGM) %>%
  summarise(n = n()) %>% 
  mutate(PercentAssisted = n / sum(n)) %>% ungroup()


threes_made %>% 
  filter(AssistedFGM == "Solo") %>% 
  arrange(desc(n)) %>% head(10)



percent_assisted <- shooting_analysis %>% 
  filter(AssistedFGM != "None", Alert == "OK") %>% 
  group_by(EventPlayerID, AssistedFGM) %>% 
  summarise(n_FGM = n()) %>% 
  mutate(PercentAssisted = n_FGM / sum(n_FGM)) %>% 
  group_by(EventPlayerID) %>% 
  mutate(total_made = sum(n_FGM)) %>% 
  left_join(players_2019 %>% select(PlayerID, PlayerName), by = c("EventPlayerID" = "PlayerID")) 

percent_assisted %>% 
  filter(AssistedFGM == "Solo",
         total_made > 50) %>% 
  arrange(desc(PercentAssisted)) %>% head(10)


percent_assisted %>% 
  filter(AssistedFGM == "Solo",
         total_made > 50) %>% 
  arrange(desc(PercentAssisted)) %>%
  ggplot(aes(x= PercentAssisted)) + geom_histogram()


percent_assisted %>% 
  filter(AssistedFGM == "Solo", total_made > 50) %>% 
  arrange(desc(PercentAssisted)) %>%
  ggplot(aes(x= PercentAssisted, y= total_made)) + geom_point() +geom_smooth(method = "lm")


percent_assisted %>% 
  distinct(EventPlayerID, .keep_all = T) %>% 
  ggplot(aes(x= total_made)) + geom_histogram()


percent_assisted %>% 
  filter(str_detect(PlayerName, "MORANT"))

percent_assisted %>% 
  filter(AssistedFGM == "Solo") %>% 
  filter(PercentAssisted >= 0.75 & total_made >200)








different_shot_types <- shooting_analysis %>% 
  filter(AssistedFGM != "None") %>%
  group_by(EventPlayerID, EventType, AssistedFGM) %>% 
  summarise(n_shots = n()) %>%
  mutate(proportion = n_shots / sum(n_shots)) %>% 
  left_join(players_2019 %>% select(PlayerID, PlayerName), by = c("EventPlayerID" = "PlayerID")) 




different_shot_types %>% 
  filter(EventType == "made3_jump") %>% 
  arrange(desc(n_shots)) %>% View()


```


